The imprint of global magma oceans on exoplanet demographics

### 2023-04-14 src/tex/ms.tex
address comments on Discussion

Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp threshold instellation~\citep{2010ppc..book.....P,2013ApJ...765..131K}.
Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone.
Its prevalent definition refers to the possibility of liquid water on an Earth-like planet with an oxidized CO$_2$/H$_2$O/N$_2$-rich atmospheres~\citep[][]{Kasting1993,Kopparapu2013,Kopparapu2014}, but it can be influenced by different internal redox states and volatile compositions~\citep{2011ApJ...734L..13P,2017ApJ...837L...4R,2018ApJ...858...72R,2019ApJ...875...31K,2020ApJ...896..115G,2022JGRE..12707456G,2023ApJ...942L..20H}.
The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1858,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in NASA's Astrophysics Data System returned $\sim\num{1700}$ results.}.
However, it should be emphasized that the habitable zone, as it stands today, is solely a concept based on theoretical predictions, and its validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017,Tuchow2023}.
The question naturally emerges if a planetary habitable zone -- in its common form with boundaries simply defined by the received stellar irradiation -- is a predictive theoretical concept and how the habitable zone hypothesis can be tested observationally.

@@

\subsection{Mission design trades}\label{sec:mission-design-trades}
To explore the impact of mission trade-offs on the detectability of the runaway greenhouse transition, we simulated different survey designs and strategies and measured their capability to recover the trend and constrain its parameters.
We assessed this capability based on two determinants: the likelihood that the mission is able to detect the injected trend, and the precision with which it can constrain the parameters of that trend.

\subsubsection{The value of follow-up campaigns}
The dependencies of a detection change when additional information beyond planet radii is available for the characterized planet population.
As runaway greenhouse phases leave a stronger imprint in bulk density than in planet radius (see Sect.~\ref{sec:res_followup}), it would be beneficial to obtain constraints on planetary masses and test the runaway greenhouse hypothesis in density space instead of radius space.
This way, useful results can be obtained under more pessimistic conditions, e.g., a low predominant water content of planetary surfaces and atmospheres or a smaller available planet sample.
For a mission design similar to PLATO, a density-based hypothesis test on about a third of the overall sample is equivalent to a pure radius-based analysis.
At a fixed sample size, key parameters of the runaway greenhouse models can be more narrowly constrained when additional mass measurements are available.

Precise ground-based radial velocity measurements will be needed to provide these data, and a number of instruments are already successfully employed in characterizing terrestrial-sized exoplanets~\citep[e.g.,][]{Queloz2001a,Pepe2010,Johnson2010b,Ribas2023}.
Recently, the National Aeronautics and Space Administration (NASA) and the National Science Foundation (NSF) commissioned an ``Extreme Precision Radial Velocity (EPRV) Initiative''~\citep{Crass2021} to develop methods and facilities for precise mass measurements of temperate terrestrial planets.
Their findings highlight that such measurements are costly, and therefore follow-up efforts may only be available for a subsample of the targets of a mission of \plato's scale.
The diagnostic power of the hypothesis tests we demonstrated here may be improved by simultaneously fitting for the trend in radius space in the subsample without RV follow-up.
An optimized mission in search for the inner edge of the habitable zone will further enhance its information content via a wise selection of follow-up targets, i.e., balancing objects located on either side of the expected instellation threshold $S_\mathrm{thresh}$.

\subsubsection{The importance of M~dwarfs in the target list}
To date, the majority of planets with radius measurements orbit FGK~dwarfs, and, based on the instellation they receive, most of them lie in the runaway greenhouse regime~\citep{Thompson2018}.
Obviously, a step-function-like signal in the exoplanet demographics like the one investigated above can not be constrained well if only one side of the discontinuity is being sampled.
This, however, is the situation for planetary systems around Sun-like stars -- their habitable zones are so distant that transiting planets within them are very rare due to pure geometrical reasons.
It was, among other reasons, the sharp drop in transit probability with orbital distance that has prompted a number of recent transit surveys to specifically target M~dwarfs~\citep[e.g.,][]{Irwin2009,Obermeier2016,Delrez2018,Sebastian2021,Dietrich2023}, but the sample of terrestrial planets orbiting them is still small~\citep[e.g.,][]{Berger2020,Hardegree-Ullman2020a}.

M~dwarf systems are also key for detecting the runaway greenhouse transition:
Our calculations with different spectral types (Sect.~\ref{sec:results_FGK_M}) show that the information content of M~dwarfs in a sample dominates the hypothesis tests.
Besides their large number in a volume/magnitude-limited sample, transiting M~dwarf planets are more likely to be located near the threshold instellation and in particular on orbits further out, i.e., in the optimistic habitable zone.
In fact, we showed that the FGK part of the planet sample barely contributes to the statistical power.
Furthermore, the transit depth difference at the transition is expected to be larger for M~dwarfs~\citep[$\sim \SI{100}{\ppm}$ for early, $\sim \SI{1000}{\ppm}$ for late M~stars,][]{Turbet2019}, enhancing the demographic signal it leaves.
An additional advantage of targeting M~dwarfs are their extended runaway greenhouse phases, whose duration can reach the order of \SI{100}{\mega\year}~\citep{Luger2015}.
This increases the probability of observing any given planet in the sample during the runaway greenhouse phase, essentially driving $f_\mathrm{rgh}$ to higher values.
Therefore, in addition to the high scientific value of boosted detections of potentially habitable planets, M dwarfs are also indispensable for the discovery and characterization of the runaway greenhouse transition.
As with a pure volume-limited sample, a targeted M~dwarf survey, too, profits from follow-up measurements of planetary masses with an order of magnitude increase in evidence.

\subsection{Constraining planetary habitability}\label{sec:habitability}
%\todo[inline]{Magma oceans influence the amount of water available at the surface and atmosphere, which is 1. commonly used as environmental marker to assess habitability, and 2. influences the planet's climate (greenhouse effect).}
A potential for liquid water on the surface of a planet is commonly used as an environmental marker to assess its habitability~\citep{Huang1959,Hart1978,Kasting1993,Kaltenegger2011,Kopparapu2013}.
The runaway greenhouse transition represents an upper bound on received irradiation for this condition.
Its detection would thus not only empirically confirm the habitable zone concept but also help to locate it in the observationally available planetary parameter space.
In Sect.~\ref{sec:res_constrain-S_thresh}, we show that the threshold instellation at which the runaway greenhouse transition occurs can be reasonably constrained without imposing overly optimistic conditions on the underlying planet population, instrumentation, or survey strategy.
A mission like \plato\ is well equipped to perform this measurement; the chances for success are directly proportional to the proportion of characterized planets around M dwarfs and to the number of planets for which masses can be determined.

The situation is different for the planetary water inventory and the fraction of planets with runaway greenhouse climates:
Since these parameters are degenerate, they cannot be well constrained without independent measurements.
This degeneracy could be lifted if independent measurements of atmospheric compositions can be made.
For example, detections of water vapor in planets above the threshold instellation, combined with precise radius measurements, would constrain the predominant water content of terrestrial planets.
A follow-up campaign with the \ariel\ mission of \plato\ planets could be considered to make this contribution.

@@

\subsubsection{Structures in the planet occurrence rate density}
In principle, the baseline occurrence rate density in radius-period space that governs the generation of our synthetic planet sample might influence our findings, especially if it contains any features that coincide with the injected demographic feature.
This is not the case in the model from \citet{Bergsten2022} that we adopted: Its occurrence rate density varies smoothly in the domain relevant for the runaway greenhouse hypothesis; transitions only occur at smaller instellations ($< \SI{50}{\watt\per\meter\squared}$) and larger radii ($> 1.6\,R_\oplus$).
We thus do not expect the model underlying our planet sample to affect the recovery of the transition.
However, it should be kept in mind that any unknown sharp features in the true distribution of terrestrial planets could complicate a detection or lead to a false positive result (see Sect.~\ref{sec:dis_falsepositive}).

@@

The models in this study do not into account a predicted additional radius increase on the percent level due to a molten interior~\citep{Bower2019}, and we neglect any potential effects stemming from a deviating gas exchange between atmosphere and interior in super-runaway greenhouse planets due to different redox conditions~\citep{Ikoma2018,2021JGRE..12606711L,2022PSJ.....3...93B,2021SSRv..217...22G}.
Future theoretical work that self-consistently models interior-atmosphere interactions is needed to take these effects into account~\citep{2022arXiv220310023L,Wordsworth2022}.

@@

\subsubsection{``Sharpness'' of the demographic trend}
The real counterpart to our idealized step-function central to our hypothesis tests could be blurred by several factors.
Its amplitude is affected by ``sub-Neptunes'' with significant Hydrogen/Helium envelopes, which can disguise as inflated rocky planets.
They might not contribute to the demographic signal we are looking for, although a similar transition was suggested for water-dominated sub-Neptunes~\citep{Pierrehumbert2022}.
A subset of such gas-rich planets will experience atmospheric loss via photoevaporation~\citep{Owen2013} or core-powered mass loss~\citep{Ginzburg2018}, turning them from planets appearing inflated into much smaller \mbox{(super-)Earths}.
Further dilution comes from the population of planets that lack an atmosphere or sufficient volatiles that can create a steam atmosphere.
There is also a time dependence, both because stellar luminosities change with age and because runaway greenhouse planets with limited water reservoirs ultimately desiccate, ending their steam atmosphere phase.

@@

All of the above mechanisms dilute the demographic imprint of the runaway greenhouse transition with contributions that are challenging to quantify.
The ``dilution factor" we introduced embodies our ignorance of these contributions.
Importantly, while this may offset our estimates of the signal's amplitude, its shape remains the same.

>>>>

Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp threshold instellation~\citep{2010ppc..book.....P,2013ApJ...765..131K}.
Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone.
Its prevalent definition refers to the possibility of liquid water on an Earth-like planet with an oxidized CO$_2$/H$_2$O/N$_2$-rich atmospheres~\citep[][]{Kasting1993,Kopparapu2013,Kopparapu2014}, but it can be influenced by different internal redox states and volatile compositions~\citep{2011ApJ...734L..13P,2017ApJ...837L...4R,2018ApJ...858...72R,2019ApJ...875...31K,2020ApJ...896..115G,2022JGRE..12707456G,2023ApJ...942L..20H}.
The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1858,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in the National Aeronautics and Space Administration (NASA) Astrophysics Data System returned $\sim\num{1700}$ results.}.
However, it should be emphasized that the habitable zone, as it stands today, is solely a concept based on theoretical predictions, and its validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017,Tuchow2023}.
The question naturally emerges if a planetary habitable zone -- in its common form with boundaries simply defined by the received stellar irradiation -- is a predictive theoretical concept and how the habitable zone hypothesis can be tested observationally.

@@

\subsection{Mission design trade studies}\label{sec:mission-design-trades}
To explore the impact of mission trade-offs on the detectability of the runaway greenhouse transition, we simulated different survey designs and strategies and measured their capability to recover the trend and constrain its parameters.
We assessed this capability based on two determinants: the likelihood that the mission is able to detect the injected trend, and the precision with which it can constrain the parameters of that trend.

\subsubsection{The value of follow-up campaigns}
The constraining power changes when additional information beyond planet radii is available for the characterized planet population.
As runaway greenhouse phases leave a stronger imprint on bulk density than on planet radius (see Sect.~\ref{sec:res_followup}), it would be beneficial to obtain constraints on planetary masses and test the runaway greenhouse hypothesis in density space instead of radius space.
This way, useful results can be obtained under more pessimistic conditions, e.g., a low predominant water content of planetary surfaces and atmospheres or a smaller available planet sample.
For a mission design similar to PLATO, a density-based hypothesis test on about a third of the overall sample is equivalent to a pure radius-based analysis.
At a fixed sample size, key parameters of the runaway greenhouse models can be more narrowly constrained when additional mass measurements are available.

Precise ground-based radial velocity measurements will be needed to provide these data, and a number of instruments are already successfully employed in characterizing terrestrial-sized exoplanets~\citep[e.g.,][]{Queloz2001a,Pepe2010,Johnson2010b,Ribas2023}.
A new generation of instruments on extremely large telescopes such as \gclef\ on the \gmt ~\citep{Szentgyorgyi2016}, \andes\ on the \elt ~\citep{Marcantonio2022}, or \modhis\ on the \tmt ~\citep{Mawet2019} will open up the discovery space even further.

Recently, NASA and the National Science Foundation (NSF) commissioned an ``Extreme Precision Radial Velocity Initiative''~\citep{Crass2021} to develop methods and facilities for precise mass measurements of temperate terrestrial planets.
Their findings highlight that such measurements are costly, and therefore follow-up efforts may only be available for a subsample of the targets of a mission of \plato's scale.
The diagnostic power of the hypothesis tests we demonstrated here may be improved by simultaneously fitting for the habitable zone inner edge discontinuity in the subsample without RV follow-up.
An optimized mission in search for the inner edge of the habitable zone will further enhance its information content via a informed selection of follow-up targets, i.e., balancing objects located on either side of the expected instellation threshold $S_\mathrm{thresh}$.

\subsubsection{The importance of M~dwarfs in the target list}
To date, the majority of planets with radius measurements orbit FGK~dwarfs, and, based on the instellation they receive, most of them lie in the runaway greenhouse regime~\citep{Thompson2018}.
Obviously, a radius/density discontinuity in the exoplanet demographics like the habitable zone inner edge discontinuity cannot be constrained well if only one side of the discontinuity is being sampled.
This, however, is the situation for planetary systems around Sun-like stars -- their habitable zones are so distant that transiting planets within them are very rare due to pure geometrical reasons.
It was, among other reasons, the sharp drop in transit probability with orbital distance that has prompted a number of recent transit surveys to specifically target M~dwarfs~\citep[e.g.,][]{Irwin2009,Obermeier2016,Delrez2018,Sebastian2021,Dietrich2023}, but the sample of terrestrial planets orbiting them is still small~\citep[e.g.,][]{Berger2020,Hardegree-Ullman2020a}.

M~dwarf systems are also key for detecting the runaway greenhouse transition:
Our calculations with different spectral types (Sect.~\ref{sec:results_FGK_M}) show that the information content of M~dwarfs in a sample dominates the hypothesis tests.
Besides their large number in a volume/magnitude-limited sample, transiting M~dwarf planets are more likely to be located near the threshold instellation and in particular on orbits further out, i.e., in the optimistic habitable zone.
In fact, we showed that the FGK part of the planet sample barely contributes to the statistical power.
Furthermore, the transit depth difference at the transition is expected to be larger for M~dwarfs~\citep[$\sim \SI{100}{\ppm}$ for early, $\sim \SI{1000}{\ppm}$ for late M~stars,][]{Turbet2019}, enhancing the demographic signal it leaves.
An additional advantage of targeting M~dwarfs are their extended runaway greenhouse phases, whose duration can reach the order of gigayears~\citep{Luger2015}.
This increases the probability of observing any given planet in the sample during the runaway greenhouse phase, essentially driving $f_\mathrm{rgh}$ to higher values.
Therefore, in addition to the high scientific value of boosted detections of potentially habitable planets, M dwarfs are also indispensable for the discovery and characterization of the runaway greenhouse transition.
As with a pure volume-limited sample, a targeted M~dwarf survey, too, profits from follow-up measurements of planetary masses with an order of magnitude increase in evidence.

\subsection{Constraining planetary habitability}\label{sec:habitability}
%\todo[inline]{Magma oceans influence the amount of water available at the surface and atmosphere, which is 1. commonly used as environmental marker to assess habitability, and 2. influences the planet's climate (greenhouse effect).}
A potential for liquid water on the surface of a planet is commonly used as an environmental marker to assess its surface habitability~\citep{Huang1959,Hart1978,Kasting1993,Kaltenegger2011,Kopparapu2013}.
The runaway greenhouse transition represents an upper bound on received irradiation for this condition.
Its detection would thus not only empirically confirm the habitable zone concept but also help to locate it in the observationally available planetary parameter space.
In Sect.~\ref{sec:res_constrain-S_thresh}, we show that the threshold instellation at which the runaway greenhouse transition occurs can be reasonably constrained without imposing overly optimistic conditions on the underlying planet population, instrumentation, or survey strategy.
A mission like \plato\ is well equipped to perform this measurement; the constraining power is directly proportional to the proportion of characterized planets around M dwarfs and to the number of planets for which masses can be determined.

The situation is different for the planetary water inventory and the fraction of planets with runaway greenhouse climates:
Since these parameters are degenerate, they cannot be well constrained without independent measurements.
This degeneracy could be lifted if independent measurements of atmospheric compositions can be made.
For example, detections of water vapor in planets above the threshold instellation, combined with precise radius measurements, would constrain the predominant water content of terrestrial planets.
%A follow-up campaign with the \ariel\ mission of \plato\ planets could be considered to make this contribution.

@@

\subsubsection{Structures in the planet occurrence rate density}
In principle, the baseline occurrence rate density in radius-period space that governs the generation of our synthetic planet sample might influence our findings, especially if it contains any features that coincide with the injected demographic feature.
This is not the case in the model from \citet{Bergsten2022} that we adopted: Its occurrence rate density varies smoothly in the domain relevant for the runaway greenhouse hypothesis; transitions only occur at smaller instellations ($< \SI{50}{\watt\per\meter\squared}$) and larger radii ($> 1.6\,R_\oplus$).
We thus do not expect the model underlying our planet sample to affect the investigation of the habitable zone inner edge discontinuity.
However, it should be kept in mind that any unknown sharp features in the true distribution of terrestrial planets could complicate a detection or lead to a false positive result (see Sect.~\ref{sec:dis_falsepositive}).

@@

The models in this study do not take into account a predicted additional radius increase on the percent level due to a molten interior~\citep{Bower2019}, and we neglect any potential effects stemming from a deviating gas exchange between atmosphere and interior in super-runaway greenhouse planets due to different redox conditions~\citep{Ikoma2018,2021JGRE..12606711L,2022PSJ.....3...93B,2021SSRv..217...22G}.
Future theoretical work that self-consistently models interior-atmosphere interactions is needed to take these effects into account~\citep{2022arXiv220310023L,Wordsworth2022}.

@@

\subsubsection{``Sharpness'' of the habitable zone inner edge discontinuity}
The real counterpart to our idealized step-function central to our hypothesis tests could be blurred by several factors.
Its amplitude is affected by ``sub-Neptunes'' with significant Hydrogen/Helium envelopes, which can disguise as inflated rocky planets.
They might not contribute to the habitable zone inner edge discontinuity, although a similar transition was suggested for water-dominated sub-Neptunes~\citep{Pierrehumbert2022}.
A subset of such gas-rich planets will experience atmospheric loss via photoevaporation~\citep{Owen2013} or core-powered mass loss~\citep{Ginzburg2018}, turning them from planets appearing inflated into much smaller \mbox{(super-)Earths}.
Further dilution comes from the population of planets that lack an atmosphere or sufficient volatiles that can create a steam atmosphere.
There is also a time dependence, both because stellar luminosities change with age and because runaway greenhouse planets with limited water reservoirs ultimately desiccate, ending their steam atmosphere phase.

@@

All of the above mechanisms may dilute the demographic imprint of the runaway greenhouse transition with contributions that are challenging to quantify.
The ``dilution factor" we introduced embodies our ignorance of these contributions.
Importantly, while this may offset our estimates of the signal's amplitude, its shape remains the same.
